---
title: "Lab_09"
author: "Hadia Beshir"
format: pdf
---

> Import Data

```{r}
candy_file <- "candy-data.csv"
candy = read.csv(candy_file, row.names=1)
head(candy)
```

**> Q1. How many different candy types are in this dataset?**

```{r}
nrow(candy)
```

According tot he function there are 85 different candy types in this data set.

**> Q2. How many fruity candy types are in the dataset?**

```{r}
sum(candy$fruity)
```

According to the function there is 38 different fruity candy types in this dataset.

We can find the winpercent value for Twix by using its name to access the corresponding row of the dataset. This is because the dataset has each candy name as rownames (recall that we set this when we imported the original CSV file). For example the code for Twix is:

```{r}
candy["Twix", ]$winpercent
```

```{r}
library(dplyr)
candy |> 
  filter(row.names(candy)=="Twix") |> 
  select(winpercent)
```

**> Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?**

```{r}
candy |> 
  filter(row.names(candy)== "Reese's Peanut Butter cup") |> 
  select(winpercent)
```

The winpercent value for Rees's is 84.18.

**> Q4. What is the winpercent value for “Kit Kat”?**

```{r}
candy |> 
  filter(row.names(candy)== "Kit Kat") |> 
  select(winpercent)
```

The winpercent value for Kit Kat is 76.77

**> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?**

```{r}
candy |> 
  filter(row.names(candy)== "Tootsie Roll Snack Bars") |> 
  select(winpercent)
```

The winpercent value for Tootsie roll is 49.65

> There is a useful skim() function in the skimr package that can help give you a quick overview of a given dataset. Let’s install this package and try it on our candy data.

```{r}
library("skimr")
skim(candy)
```

**> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?**

```{r}
summary(candy)
```


**>Q7. What do you think a zero and one represent for the candy$chocolate column?**

```{r}
table(candy$chocolate)
```

A zero in the candy chocolate coloumn shows that there are that number of candys that do not contain chocolate (false), and a 1 means that there are that amount of candys that do contain chocolate. In this case, 48 cnadys dont have chocolate and 37 do.

**> Q8. Plot a histogram of winpercent values using both base R an ggplot2.**

>Plot using R

```{r}
hist(candy$winpercent)
```

> Plot using ggplot2

```{r}
library(ggplot2)
ggplot(candy) + 
  aes(winpercent) + 
  geom_histogram(bins=10, fill="blue", col="orange")
```

**> Q9. Is the distribution of winpercent values symmetrical?**

It is clear that the distribution is slightly skewed to the right. There are two peak areas in the data. Despite the two peaks the data is relativley packed in the middle.

**> Q10. Is the center of the distribution above or below 50%?** 

```{r}
median(candy$winpercent)
mean(candy$winpercent)
```

According to the functions ran above, the center of distribution (better explained by the median) is below 50%.

**> Q11. On average is chocolate candy higher or lower ranked than fruit candy?**

```{r}
mean(candy$winpercent[candy$chocolate == 1])
```

```{r}
mean(candy$winpercent[candy$fruity == 1])
```

According to the data, chocolate has a higher average than fruity does. This suggests that chocolate is higher ranked.

**> Q12. Is this difference statistically significant?**

```{r}
t.test(candy$winpercent[as.logical(candy$chocolate)], 
       candy$winpercent[as.logical(candy$fruity)])
```

The p-value, as provided above, is 2.871e-08. If the p-value is less than 0.05, then the result is significant. Since the p-value is far below 0.05, the results have high statistical significance.

**> Q13. What are the five least liked candy types in this set?** 

```{r}
head(candy[order(candy$winpercent), ], n=5)
candy %>% arrange(winpercent) %>% head(5)
```
The five least liked candy types in this set: 
1. Nik L Nip	
2. Boston Baked Beans		
3. Chiclets
4. Super Bubble	
5. Jawbusters

**> Q14. What are the top 5 all time favorite candy types out of this set?**

```{r}
head(candy[order(candy$winpercent, decreasing = TRUE), ], n=5)
candy %>% arrange(desc(winpercent)) %>% head(5)
```
The top 5 all time favorite candy types out of this set are: 
1. Reese's Peanut Butter cup		
2. Reese's Miniatures	
3. Twix
4. Kit Kat
5. Snickers

**> Q15. Make a first barplot of candy ranking based on winpercent values.**

```{r, fig.height=8, fig.width=5}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

**> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?**

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col() +
  theme(axis.text.y = element_text(size = 2)) 
```

> Let’s setup a color vector (that signifies candy type) that we can then use for some future plots. We start by making a vector of all black values (one for each candy). Then we overwrite chocolate (for chocolate candy), brown (for candy bars) and red (for fruity candy) values.

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) +
  theme(axis.text.y = element_text(size = 2))
```

**> Q17. What is the worst ranked chocolate candy?**

The worst ranked chocolate candy is Reeses peanut Butter cup.

**> Q18. What is the best ranked fruity candy?**

The best ranked frity candy is Starburst.

Taking a look at pricepercent

> To this plot we will add text labels so we can more easily identify a given candy. There is a regular geom_label() that comes with ggplot2. However, as there are quite a few candies in our dataset lots of these labels will be overlapping and hard to read. To help with this we can use the geom_text_repel() function from the ggrepel package.

```{r}
library(ggrepel)
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)

```

**>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?**

```{r}
head(candy[order(candy$pricepercent), c("pricepercent", "winpercent")], n=5)
```

Tootsie Roll Midgies is the cheapest candy and has the most bank for your buck.

**> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?**

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

The most expensive candy and the least popular are: 
1. Nik L Nip	
2. Nestle Smarties	
3. Ring pop	
4. Hershey's Krackel	
5. Hershey's Milk Chocolate

Exploring the correlation structure

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

**> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?**
The two most aniti correlated variables is fruity and chocolate.

**> Q23. Similarly, what two variables are most positively correlated?**

The two most positively correlated variables are winipercent and chocolate.

Principal Component Analysis

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

> We can make a much nicer plot with the ggplot2 package but it is important to note that ggplot works best when you supply an input data.frame that includes a separate column for each of the aesthetics you would like displayed in your final plot. 

```{r}
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)
p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

If you want to see more candy labels you can change the max.overlaps value to allow more overlapping labels or pass the ggplot object p to plotly like so to generate an interactive plot that you can mouse over to see labels:
```{r}
library(plotly)
ggplotly(p)
```

Let’s finish by taking a quick look at PCA our loadings. Do these make sense to you? Notice the opposite effects of chocolate and fruity and the similar effects of chocolate and bar (i.e. we already know they are correlated).

```{r}
res <- as.data.frame(pca$rotation)
ggplot(res) +
  aes(PC1, reorder(rownames(res), PC1)) +
  geom_col()
```

**>Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?**

The variables that have the most positive direction and point to the right are fruity, pluribus, and hard. This correlation has been seen in the plots above, especially the anti correlation plot where fruity area was on the right while chcoolate was on the left. I would say this does make sense because fruity candys inclduing gummies, sourpatch, skittles, are all coming in small packs and are slightly harder an more chewy.

**> Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?**

Based on the plots we generated so far and my udnerstanding of  the test is that people like chocolate (mainly chocolates that have biscuits or peanut flavor in it) the most and are willing to spend their money on it. Many are okay with buying cheaper fruity candies, hwoevver, as soon as the price goes up, people are not likely to buy.