---
title: "Lab_08"
author: "Hadia Beshir"
format: pdf
---

> 1. Preparing the Data

```{r}
fna.data <- "WisconsinCancer.csv"
wisc.df <- read.csv(fna.data, row.names=1)
head(wisc.df, 4)
```
A data frame that omits the first coloumn:
```{r}
wisc.data <- wisc.df[,-1]
```

```{r}
# Create diagnosis vector for later
diagnosis <- as.factor(wisc.df$diagnosis)
```

>**Q1. How many observations are in this dataset?**

```{r}
nrow(wisc.df)
```
There are 569 observations in the dataset.

>**Q2. How many of the observations have a malignant diagnosis?**

```{r}
table(wisc.df$diagnosis)
```

212 of the observations have a malignant analysis while 357 have a benign analysis.

>**Q3. How many variables/features in the data are suffixed with _mean?**

```{r}
length(grep("_mean", colnames(wisc.df)))
```
10 variables/features have _mean.

Principal Component Analysis.
```{r}
# Check column means and standard deviations
colMeans(wisc.data)

apply(wisc.data,2,sd)
```
```{r}
# Perform PCA on wisc.data through the following code
wisc.pr <- prcomp(wisc.data, scale = TRUE)
```
```{r}
# Look at summary of results
summary(wisc.pr)
```

>**Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?**

```{r}
summary(wisc.pr)
```
When looking at the summary, it is clear that the first principal component had 44.27% of the original variance. 

>**Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?**

```{r}
pr.var <- wisc.pr$sdev^2
pve <- pr.var / sum(pr.var)
cumulative_pve <- cumsum(pve)
which(cumulative_pve >= 0.70)[1]
```
According to the results above, 3 principal compnents are needed to describe at least 70% of the original variance in the data.

>**Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?**

```{r}
pr.var <- wisc.pr$sdev^2
pve <- pr.var / sum(pr.var)
cumulative_pve <- cumsum(pve)
which(cumulative_pve >= 0.90)[1]
```
According to the results above, 7 principal compnents are needed to describe at least 90% of the original variance in the data.

Interpreting PCA Results

>**Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?**

```{r}
biplot(wisc.pr)
```
The plot generated is evry hard to udnerstand. It is clear that it is overcrowded and nothing can be easily red. This may be due to the fact that the data is large, this is not a small dataset.
```{r}
# Scatter plot observations by components 1 and 2
library(ggplot2)
df <- as.data.frame(wisc.pr$x)
ggplot(df) +
  aes(x=PC1, y=PC2, col=diagnosis) +
  geom_point()
```

>**Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?**

```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC3, col=diagnosis) +
  geom_point()
```
PC3 explains less of the variance than PC2 which causes less distinction in data and more verlapping.

Variance Explained
```{r}
# Calculate variance of each component
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

```{r}
# Variance explained by each principal component: pve
pve <- pr.var / sum(pr.var)

# Plot variance explained for each principal component
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Percent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```
```{r}
## ggplot based graph
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```
Communicating PCA results

>**Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?**

```{r}
wisc.pr$rotation["concave.points_mean", 1]
sort(abs(wisc.pr$rotation[,1]), decreasing = TRUE)
```
The results for concave.points_mean came out to be -0.2608522. There most likely are other features like the mean that may show higher contributions to PC1.

Hierarchical Clsutering
```{r}
# Scale the wisc.data data using the "scale()" function
data.scaled <- scale(wisc.data)
```
```{r}
data.dist <- dist(data.scaled)
```
```{r}
wisc.hclust <- hclust(data.dist, method = "complete")
```

Results of Hierarchical Clustering

>**Q10. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?**

```{r}
plot(wisc.hclust)
abline(h=20, col="pink", lty=2)
```
The height at which the clustering model has 4 clusters is around 19.

Selecting Number of Clusters
```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k = 4)
```

```{r}
table(cutree(wisc.hclust, k = 4), diagnosis)
```
>**Q11. OPTIONAL: Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 6? How do you judge the quality of your result in each case?**

```{r}
table(cutree(wisc.hclust, k = 6), diagnosis)
```
After increasin the k value to 6, the maligant tumors data began to spread out an split into sub clusters. This doesnt mean there is better matches, it just is spliting the data. Outliers are a better way to determine the "better data".

>**Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.**

```{r}
wisc.hclust.complete <- hclust(data.dist, method="complete")
wisc.hclust.average  <- hclust(data.dist, method="average")
wisc.hclust.single   <- hclust(data.dist, method="single")
wisc.hclust.ward     <- hclust(data.dist, method="ward.D2")
View(wisc.df)
```

```{r}
table(cutree(wisc.hclust.complete, k=4), diagnosis)
table(cutree(wisc.hclust.ward, k=4), diagnosis)
```
The method that gives my favorite results for the samedata set is Ward.D2. This is becayse it seems to have higehr accuracy according to the results. it also has the most of one type per a cluster which can indicate its more pure.

Combining Methods
```{r}
wisc.pr.hclust <- hclust(dist(wisc.pr$x[, 1:7]), method="ward.D2")
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
table(grps, diagnosis)
```
```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC2) +
  geom_point(col=grps)
```
```{r}
library(rgl)
plot3d(wisc.pr$x[,1:3], xlab="PC 1", ylab="PC 2", zlab="PC 3", cex=1.5, size=1, type="s", col=grps)
```
```{r}
## Use the distance along the first 7 PCs for clustering i.e. wisc.pr$x[, 1:7]
wisc.pr.hclust <- hclust(dist(wisc.pr$x[, 1:7]), method="ward.D2")
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=2)
```

>**Q13. How well does the newly created hclust model with two clusters separate out the two “M” and “B” diagnoses?**

```{r}
table(grps, diagnosis)
```
The model helps to seperae the diagnoses very well The clustering in the original since it had too much cluter. PCA takes away all the extra which allows us to view the cluster and structure.

>**Q14. How well do the hierarchical clustering models you created in the previous sections (i.e. without first doing PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.hclust.clusters and wisc.pr.hclust.clusters) with the vector containing the actual diagnoses.**

```{r}
table(wisc.hclust.clusters, diagnosis)
table(grps, diagnosis)
```
It is clear that the PCA based model is better as it reduced the data to only the first couple cs which gets rid fo all the extra clutter which allowes for better grouping and struture .

> 6. Sensitivity/specifity

>**Q15. OPTIONAL: Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?**

It is clear that the PCA based hierarchical model was better for the balance of both. The PCA aggregatpr helps to eliminate all the extra nois. the wards linkage helps to identify clusters that helps us better understand the diagnosis.

> 7. Prediction

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```
```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

>**Q16. Which of these new patients should we prioritize for follow up based on your results?**

Based on the plot data, patient 1 should be prioritized fro a follow up. This is because patient 1 has points that are black/red which can be identifies as malignant. Patient 2 had points that primarily showed bengin.

> 8. About this document.

A key takeaway is that preprocessing with PCA before clustering often yields better results than clustering on raw data alone. This workflow (dimensionality reduction followed by clustering) is widely used in biomedical research for tasks like patient stratification, biomarker discovery, and disease subtyping.
```{r}
sessionInfo()
```

