---
title: "Lab07"
author: "Hadia Beshir"
format: pdf
---

**In class lesson**

> Generate some data

```{r}
tmp_x <- c(rnorm(15, -2), rnorm(15, 2))
tmp_y <- c(rnorm(15, 2), rnorm(15, -2))
df <- cbind(x=tmp_x, y=tmp_y)
```

> Plot the data

```{r}
plot(df)
```

> Lets do a k-means clustering analysis

```{r}
km <- kmeans(df, 2, nstart=20)
km
```

```{r}
attributes(km)
```

> Q: Number of points per a cluster ?

```{r}
km$size
```
> Q: cluster size ?
above
> Q: Cluster assignment?

```{r}
km$cluster
```
```{r}
table(km$cluster)
```
> Q: Cluster center ?

```{r}
km$centers
```

> Plot kmeans clustering

```{r}
plot(df, col=c("RED", "BLACK")[km$cluster])
points(km$centers, col="BLUE", pch="X")
```
Plot with ggplot:
```{r}
library(ggplot2)
ggplot(df) + 
  aes(x, y, col=factor("km$cluster")) + 
  geom_point() +
  scale_color_manual(values=c("RED", "BLACK"))
geom_point(data=km$centers, col="BLUE", shape="X", size=3, aes(x, y))
```
> Demonstrate the use of dist(), hclust(), plot(), and cutree()

```{r}
dm <- dist(df)
```

```{r}
hc <- hclust(dm)
```
```{r}
plot(hc)
```

Cut tree into 2 clusters

```{r}
groups <- cutree(hc, k=2) 
print(groups)

plot(df, col=groups, main="Hclust Cluster")
```


**Lab 07**

> Data Import

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
```

>**Q: How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?**

```{r}
dim(x)
```

> Checking your data: preview the first 6 rows.

```{r}
head(x)
```
> Hmm, it looks like the row-names here were not set properly as we were expecting 4 columns (one for each of the 4 countries of the UK - not 5 as reported from the dim() function).

> Here it appears that the row-names are incorrectly set as the first column of our x data frame (rather than set as proper row-names). This is very common and sometimes what we want - but not in this case. Lets try to fix this up with the following code, which sets the rownames() to the first column and then removes the troublesome first column (with the -1 column index):

```{r}
rownames(x) <- x[,1]
x <- x[,-1]
head(x)
```
```{r}
dim(x)
```
> Side-note: An alternative approach to setting the correct row-names in this case would be to read the data file again and this time set the row.names argument of read.csv() to be the first column (i.e. use argument setting row.names=1), see below:

```{r}
x <- read.csv(url, row.names=1)
head(x)
```
>**Q2:Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?**

I would prefer the one where we read the csv. This is because it is more simple and generally more used when trying to aquire data from a file. The other approach also creates a whole data frame first, then modify it. This causes more room for mistakes.

> Supporting major differences and trends:

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```
>**Q3: Changing what optional argument in the above barplot() function results in the following plot?**

Changing the beside functon result is what causes the values to be stacked on top. In teh barplot above, beside = 2, this is why each segment is next to each other. When besid = false, the results are stacked on top of each other.

> Using the ggplot2 package for these relatively simple figures is somewhat more convoluted than using “base” R. This is due to the ggplot requirement for so-called “long” data rather than the conventional “wide” data format that we currently have. We will talk more about “long” format data a little later in the course when we introduce the dplyr package.

```{r}
dim(x)
head(x)
```
```{r}
library(tidyr)
x_long <- x |> 
          tibble::rownames_to_column("Food") |> 
          pivot_longer(cols = -Food, 
                       names_to = "Country", 
                       values_to = "Consumption")

dim(x_long)
```
```{r}
head(x_long)
```

> Create a bar plot:

```{r}
library(ggplot2)
ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col(position = "dodge") +
  theme_bw()
```
>**Q4: Changing what optional argument in the above ggplot() code results in a stacked barplot figure?**

Through default ggplot has a position arguent within its geom_col() function. To make the bars stacked on top of ecah other, you must make position = stack. When position = dodge, the bars are side by side. 

>**Q5: We can use the pairs() function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?**

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)
```
```{r}
library(pheatmap)

pheatmap( as.matrix(x) )
```

>**Q6. Based on the pairs and heatmap figures, which countries cluster together and what does this suggest about their food consumption patterns? Can you easily tell what the main differences between N. Ireland and the other countries of the UK in terms of this data-set?**

Based on the map above, it is clear that none of the countries are toof ar off rom each other. When looking at the map even more deeply, its clear that N. Ireland is slightly more distinct. This is becasue in some regions liek the frehs fruit area, the country had lower values compared to the rest. Nothern ireland seems to have had lower values in regions lke other-meat and fresh fruit. Overall the countries seem to have similar food/eating patterns.

> PCA to the rescue:

```{r}
pca <- prcomp( t(x))
summary(pca)
```

>**Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points.**

```{r}
df <- as.data.frame(pca$x)
df$Country <- rownames(df)
ggplot(pca$x) +
  aes(x = PC1, y = PC2, label = rownames(pca$x)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5) +
  xlim(-270, 500) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()
```

>**Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document.**

```{r}
df <- as.data.frame(pca$x)
df$Country <- rownames(df)
unique(df$Country)
ggplot(df) +
  aes(x = PC1, y = PC2, label = Country) +
  geom_point(size = 3) +
  geom_text(aes(col= Country), vjust = -0.5) + 
  scale_color_manual(values = c("England"="orange",
                                "Scotland"="blue", 
                                "Wales"="red", 
                                "N.Ireland"="green")) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()
```

> Below we can use the square of pca$sdev , which stands for “standard deviation”, to calculate how much variation in the original data each PC accounts for.

```{r}
v <- round( pca$sdev^2/sum(pca$sdev^2) * 100 )
v
```

```{r}
z <- summary(pca)
z$importance
```

> This information can be summarized in a plot of the variances (eigenvalues) with respect to the principal component number (eigenvector number), which is given below.

```{r}
variance_df <- data.frame(
  PC = factor(paste0("PC", 1:length(v)), levels = paste0("PC", 1:length(v))),
  Variance = v
)

ggplot(variance_df) +
  aes(x = PC, y = Variance) +
  geom_col(fill = "steelblue") +
  xlab("Principal Component") +
  ylab("Percent Variation") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0))
```

> Variable Loadings

```{r}
ggplot(pca$rotation) +
  aes(x = PC1, 
      y = reorder(rownames(pca$rotation), PC1)) +
  geom_col(fill = "steelblue") +
  xlab("PC1 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```

>**Q9: Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?**

```{r}
ggplot(pca$rotation) +
  aes(x = PC2, 
      y = reorder(rownames(pca$rotation), PC2)) +
  geom_col(fill = "blue") +
  xlab("PC2 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```

The two food groups that are frequently featured is softd rinks and fresh potatoes. PC2 helps distinguish Scotland, England, Wales, and Northern Ireland based on their food consumptions. Compared to PC1, PC2 shows the second chunk of variety.