---
title: "Lab_11"
author: "Hadia Beshir"
format: pdf
---

> 5. The EBI AlphaFold database

Querying the AlphaFold database

1. Sequence to search AFDB: PQITLWQRPLVTIKIGGQLKEALLDTGADDTVLEEMSLPGRWKPKMIGGIGGFIKVRQYDQILIEICGHKAIGTVLVGPTPVNIIGRNLLTQIGCTLNF
2. Examine one of your top hits

Protein Name: Gag-Pol polyprotein
UniProt ID: P04585
Organism: Human immunodeficiency virus type 1 (HIV-1)

3. Visit the corresponding AFDB structure page. It shows basic information about the protein (drawn from UniProt), and three separate outputs from AlphaFold.

My model is a dark blue color. This shows that my mdoel is very confident (pLDDT > 90). It also states that the average pLDDT is 96.78 which is considered very high.

After locating the C-terminus of the mdoel in this case was the last letter (F). It had a pLDDT value of 88.81 (Confident). It is still considered high, just not as high as the average. it was a light/teal blue color.

The PAE matrix is all dark green. The C-terminus is highly reliable.

> 6. Generating your own structure positions
> 7. Interpreting Results

The graph generated is dark orange/red. This shows that the structure has high confidence.

ƒigure 18.

```{r}
library(imager)
img <- load.image("Figure 18..png")
plot(img)
```


> 8. Custom Analysis

```{r}
results_dir <- "HIV_Protease_Dimer_23119" 
pdb_files <- list.files(path=results_dir,
                        pattern="*.pdb",
                        full.names = TRUE)

# Print our PDB file names
basename(pdb_files)
library(bio3d)

pdbs <- pdbaln(pdb_files, fit=TRUE, exefile="msa")

```

```{r}
pdbs
```

RMSD is a standard measure of structural distance between coordinate sets. We can use the rmsd() function to calculate the RMSD between all pairs models

```{r}
rd <- rmsd(pdbs, fit=T)
range(rd)

```

Draw a heatmap of these RMSD matrix values

```{r}
library(pheatmap)

colnames(rd) <- paste0("m",1:5)
rownames(rd) <- paste0("m",1:5)
pheatmap(rd)
```

Now lets plot the pLDDT values across all models. Recall that this information is in the B-factor column of each model and that this is stored in our aligned pdbs object as pdbs$b with a row per structure/model.

```{r}
pdb <- read.pdb("1hsg")
plotb3(pdbs$b[1,], typ="l", lwd=2, sse=pdb)
points(pdbs$b[2,], typ="l", col="red")
points(pdbs$b[3,], typ="l", col="blue")
points(pdbs$b[4,], typ="l", col="darkgreen")
points(pdbs$b[5,], typ="l", col="orange")
abline(v=100, col="gray")
```

We can improve the superposition/fitting of our models by finding the most consistent “rigid core” common across all the models. For this we will use the core.find() function:

```{r}
core <- core.find(pdbs)
core.inds <- print(core, vol=0.5)
```

```{r}
xyz <- pdbfit(pdbs, core.inds, outpath="corefit_structures")
```

Figure 19.

```{r}
library(imager)
img2 <- load.image("Figure 19..png")
plot(img2)
```


Now we can examine the RMSF between positions of the structure. RMSF is an often used measure of conformational variance along the structure:

```{r}
rf <- rmsf(xyz)

plotb3(rf, sse=pdb)
abline(v=100, col="gray", ylab="RMSF")
```

Predicted Alignment Error for domains

```{r}
library(jsonlite)

# Listing of all PAE JSON files
pae_files <- list.files(path=results_dir,
                        pattern=".*model.*\\.json",
                        full.names = TRUE)
pae1 <- read_json(pae_files[1],simplifyVector = TRUE)
pae5 <- read_json(pae_files[5],simplifyVector = TRUE)

attributes(pae1)
```

```{r}
# Per-residue pLDDT scores 
#  same as B-factor of PDB..
head(pae1$plddt)
```

The maximum PAE values are useful for ranking models. Here we can see that model 5 is much worse than model 1. The lower the PAE score the better. How about the other models, what are thir max PAE scores?

```{r}
pae1$max_pae
pae5$max_pae
```

We can plot the N by N (where N is the number of residues) PAE scores with ggplot or with functions from the Bio3D package:

```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)")
```

```{r}
plot.dmat(pae5$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

Same Z-range:

```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

Residue conservation from alignment file

```{r}
aln_file <- list.files(path=results_dir,
                       pattern=".a3m$",
                        full.names = TRUE)
aln_file
```

```{r}
aln <- read.fasta(aln_file[1], to.upper = TRUE)
```

How many sequences are in this alignment
```{r}
dim(aln$ali)

```

We can score residue conservation in the alignment with the conserv() function

```{r}
sim <- conserv(aln)
plotb3(sim[1:99], sse=trim.pdb(pdb, chain="A"),
       ylab="Conservation Score")
```

Note the conserved Active Site residues D25, T26, G27, A28. These positions will stand out if we generate a consensus sequence with a high cutoff value:

```{r}
con <- consensus(aln, cutoff = 0.9)
con$seq
```

For a final visualization of these functionally important sites we can map this conservation score to the Occupancy column of a PDB file for viewing in molecular viewer programs such as Mol*, PyMol, VMD, chimera etc.

```{r}
m1.pdb <- read.pdb(pdb_files[1])
occ <- vec2resno(c(sim[1:99], sim[1:99]), m1.pdb$atom$resno)
write.pdb(m1.pdb, o=occ, file="m1_conserv.pdb")

```

Figure 20

```{r}
img3 <- load.image("Figure 20.png")
plot(img3)
```



